---
title: "Untitled"
author: "Kimberly Hinz"
date: "2020/12/08"
output:
  word_document:
    reference_docx: Styles1.docx
    fig_caption: true
    # toc: true
    # toc_depth: 5
bibliography: References.bib
csl: IJSEM.csl
---

\newpage
# Abstract
*Mixta* is a recently described genus within the bacterial family *Erwiniaceae* that is closely related to *Pantoea*, *Erwinia*, and *Tatumella*. Depending on the genes and method used for phylogenetic analysis, the genus with which *Mixta* shares the most recent ancestor differs. This study aimed to determine the cause behind these contentious results, and identified genes that may have resulted from horizontal gene transfer. Ten complete or partial genomes – two *Mixta* species, six species from other *Erwiniaceae* genera, and two outgroups – were retrieved from NCBI and annotated. Homologous genes were extracted yielding a dataset of 799 genes. The genes were aligned With the ClustalW algorithm and MEGA-CC was used for model testing and computing phylogenetic trees for each gene for both nucleotide and amino acid sequences. Distance matrices were then extracted from these trees. Nucleotide and amino acid identity analyses were also done in `R` programming language. *Pantoea* was the closest relative to the *Mixta* species in most of the analyses; however, results were not consistent across analyses. Some genes were also found to be more similar to other, non-*Pantoea* species. Diligence must be given to the selection of genes for phylogenetic analysis and to the method chosen to prevent xenologous signal from distorting the true relationships. Furthermore, future research should consider that different phylogenetic analyses may give different results.

\newpage
# Introduction
The bacterial family Erwiniaceae includes diverse Gram-negative [@Winslow1917; @Prakash2015] organisms including the genera *Mixta*, *Pantoea*, *Erwinia*, *Tatumella*, *Buchnera*, *Wigglesworthia*, *Phaseolibacter*, *Kalamiella* (LPSN, http://www.bacterio.net/). Three of these, *Mixta*, *Pantoea*, and *Erwinia*, have been the focus of numerous studies given their relevance to plant and human pathogens. For example, the genus *Erwinia* contains plant pathogens (e.g. E. amylovora) [@HaubenSwings2005; @Lopez2011; @Matsuura2012], epiphytes (e.g. E. tasmaniensis) [@Geider2006; @Mergaert1999], species that may contribute to plant disease [@Buonaurio2015], and others that are insect-associated [@Campillo2015; @Harada1997; @Skrodenyte-Arbaciauskiene2012]. Similarly, *Pantoea*, a genus reported by Gavini et al.[@Gavini1989], includes species that inhabit a variety of environments [@Gavini1989; @Popp2010; @Liu2013], including human and plant pathogens [@Prakash2015; @deBaere2004; @Fullerton2007; @Kratz2003; @Lim2006; @Schmid2003; @vanRostenberghe2006] as well as species associated with plant growth [@Smits2011; @Kim2012], insects [@Palmer2016], and fungi [@Ma2016; @Rong2016]. *Mixta*, originally *Pantoea*, is a novel genus first suggested by Palmer et al. [@Palmer2017] and then later proposed by the same researchers in 2018 [@Palmer2018]. Two species, M. calida and M. gaviniae (previously species of *Pantoea*), were originally found in infant formula and its production environment [@Popp2010]. Fritz et al. [@Fritz2014] demonstrated that some strains of M. calida were linked with post-surgical meningitis in humans. Currently, there are only five *Mixta* species validly published (LPSN, http://www.bacterio.net/).

Understanding the phylogeny of bacterial species and genes allows researchers to investigate the evolutionary history of the species, their relatedness, and the individual genes that allow the species to survive under various conditions [@Palmer2017; @Bennett2012; @Prasanna2013; @Angus2014; @deMaayer2014; @Fouts2016]. The most common methods in phylogenetic analysis are those including 16S ribosomal RNA (rRNA) sequences [@Konstantinidis2007], multi-locus sequence analysis (MLSA) with several housekeeping genes [@Konstantinidis2007; @Gevers2005; @Glaeser2015], and ribosomal MLSA with 53 structural ribosomal proteins [@Palmer2017; @Bennett2012; @Jolley2012]. However, the phylogenetic trees generated from these methods may not be very robust [@Konstantinidis2007; @Glaeser2015; @Brady2008] because of the lower phylogenetic signal from the highly conserved housekeeping genes [@Konstantinidis2007; @Gevers2005; @Fox1992; @Konstantinidis2005; @Staley2006; @RichterRosselloMora2009]. This is particularly seen within Erwiniaceae wherein most of our understanding of the phylogeny and interrelationships are based on the 16S rRNA genes [@Francino2006; @Hauben1998; @Naum2008; @Sproer1999; @Adeolu2016] because of its ubiquity. However, it has insufficient resolution at the species level [@Glaeser2015] and the genera *Pantoea*, *Erwinia*, *Tatumella*, and *Mixta* are often not monophyletic [@Rezzonico2016]. Another issue that may arise with MLSA is unknowingly including paralogues (i.e. genes originating from an intragenomic duplication event) or xenologues (i.e. genes originating from horizontal gene transfer) [@Palmer2017; @Koonin2005].

The DNA topoisomerase (gyrase) subunit (gyrB), DNA-directed RNA polymerase subunit (rpoB), ATP synthase subunit (atpD), and initiation translation factor (infB) have been shown to be more reliable than 16S rRNA gene sequences for determining intra- and intergeneric relationships [@Hedegaard1999; @Dauga2002; @Mollet1997; @Paradis2005] and a robust method for the Erwiniaceae family [@Brady2010]. The genera within Erwiniaceae are monophyletic when concatenated partial gyrB, rpoB, atpD, and infB gene sequences are used [@Brady2008; @Brady2009; @Moretti2011].

Nevertheless, despite efforts to find a suitable method for identifying the species tree for Erwiniaceae genera, *Mixta*’s position with relation to *Pantoea*, *Erwinia*, and *Tatumella* vastly differs between studies [@Prakash2015; @Palmer2017; @Palmer2018; @Adeolu2016; @Brady2011; @Rezzonico2016; @Brady2012; @Gueule2015]. Not only do different gene combinations provide different results, but different algorithms and models for alignments and phylogenetic trees also contribute to the contentious results [@Naum2008; @Octavia2014]. For example, Prakash et al. [@Prakash2015], Palmer et al. [@Palmer2017], Rezzonico et al. [@Rezzonico2016], and Gueule et al. [@Gueule2015] all used gyrB, rpoB, atpD, and infB nucleotide gene sequences in their phylogenetic analyses. Prakash et al. conducted a neighbour-joining statistical method while the others used a maximum-likelihood (ML) method. Three of the studies – Rezzonico et al., Prakash et al., and Gueule et al. – placed the *Mixta* species within the *Pantoea* genera whereas the analysis in Palmer et al. placed *Mixta* outside of the other three genera. Palmer et al. [@Palmer2018], Rezzonico et al. [@Rezzonico2016], and Brady et al. [@Brady2012] also ran analyses with the four MLSA genes as amino acid sequences. These three studies show contentious results as Rezzonico et al. found that *Mixta* is related closest with *Tatumella*, Palmer et al. found that *Mixta* is related closest with *Pantoea*, and Brady et al. found *Mixta* within the *Pantoea* genus. In a separate study conducted by Palmer et al. @Palmer2017, two ML trees were computed, one with 1039 core amino acid sequence and the other with 52 ribosomal MLSA nucleotide genes. Both ML trees contained species from *Mixta*, *Pantoea*, *Erwinia*, and *Tatumella*; however, the analysis with the core genes placed *Pantoea* and *Tatumella* as sister taxa followed by two *Mixta* species whereas the analysis with ribosomal MLSA genes placed *Pantoea* and *Tatumella* as sister taxa followed by *Erwinia* and then by the *Mixta* species. 

There are two possible causes behind contentious results in phylogenetic analyses. One is the use of different methods, including the use of a different phylogenetic tree model from another study despite both studies conducting maximum likelihood trees and the same sequence type. Additionally, there is no common phylogenetic analysis model between nucleotide and amino acid sequences; therefore, evolutionary history is estimated differently. The other reason may be horizontal gene transfer (HGT) events within the *Erwiniaceae* or from a non-closely-related species. HGT events occur frequently within bacteria and often the genes provide the bacteria the ability to adapt to new environments and/or survive selective pressures.

The objective of this paper was to determine the cause behind the contentious phylogenetic results for the *Mixta* species. To accomplish this, homologous genes were extracted from ten complete or partial genomes - two for each of *Mixta*, *Pantoea*, *Erwinia*, and *Tatumella*, and two outgroups - that were retrieved from public databases. In this study, both the nucleotide and amino acid sequences for the same genes were separately analyzed. Model testing was done for each gene in the dataset and phylogenetic trees were computed using the recommended model. Distance matrices were exported from the phylogenetic trees and these were used to determine which species were most closely related to *Mixta* for each gene. Nucleotide and amino acid identity analyses were also conducted for each gene in order to compare the results between evolutionary history and sequence similarity. Phylogenetic trees were then done with either one gene or multiple genes to confirm results.

# Materials and Methods
## Genomes and Gene Sets
Publicly available full genomes were retrieved from NCBI comprising of ten type strains from *Mixta*, *Pantoea*, *Erwinia*, *Tatumella*, *Enterobacter*, and *Pseudomonas* (Table 1). *Enterobacter cloacae* is in the same order of *Enterobacterales* as *Mixta*, *Pantoea*, *Erwinia*, and *Tatumella* whereas *Pseudomonas syringiae* is in the same class of *Gammaproteobacteria*. The genomes were annotated with Prokka v1.14.1 [@Seemann2014] using default parameters and homologous genes were extracted using GET_HOMOLOGUES software package [@Contreras-Moreira2013] with the bidirectional best-hit search algorithm using default parameters. GET_HOMOLOGUES extracted 954 homologous genes between the ten species.

```{r echo = FALSE, message = FALSE, fig.cap=TRUE}
strain <- data.frame(Species = c("Mixta calida", "Mixta gaviniae", "Pantoea agglomerans", "Pantoea septica", "Erwinia amylovora",
                                 "Erwinia tasmaniensis", "Tatumella ptyseos", "Tatumella saanichensis", 
                                 "Enterobacter cloacae subsp cloacae", "Pseudomonas syringae pv syringae"),
                     Strain = c("DSM_22759", "DSM_22758", "NBRC_102470", "LMG_5345", "CFBP_1232", "ET1/99", "NCTC_11468",
                                "NML_06-3099", "ATCC_13047", "ICMP_3023"),
                     `GenBank Accession No.` = c("GCA_002953215.1", "GCA_002953195.1", "GCA_001598475.1", "GCA_002095575.1", "GCA_000367625.2",
                                                 "GCA_000026185.1", "GCA_900478715.1", "GCA_000439375.1", "GCA_000025565.1", "GCA_001401075.1"),
                     Level = c("complete genome", "complete genome", "contigs", "contigs", "contigs", "complete genome", "complete genome", 
                               "contigs", "complete genome", "scaffold"))

library("knitr")
kable(strain, caption = "Table 1. Genomes used in this study.")
```

In `R` [@RCore2019], two filter functions were written to pass through gene files of the nucleotide sequences. Since both nucleotide and amino acid sequences were analyzed, the filters were only applied to the nucleotide sequences. A list of the passed nucleotide gene files was used to isolate the corresponding amino acid gene files to be used in later analyses.

The first filter passed through gene files that contained ten, single copy sequences. A minimum relative length was set where all sequences in a gene file had to be at least 90% the length of the longest sequence in the same file. For example, if the longest sequence in a file is 1000 bp, then the rest of the sequences had to be at least 900 bp. Of 954 genes, 38 genes were excluded for having at least extra sequence and an additional 175 genes were excluded because at least one sequence did not meet the length requirements.

However, there may be an evolutionary reason why at least one of the sequences did not meet the length requirements. The second filter took the 175 gene files that were denied by the first filter one at a time and split the sequences into two groups; the first group contained the gene sequences that were over the 90% length cutoff and the second contained the rest that were shorter than the cutoff length. The first requirement for this filter was that all sequences in the file were longer than 80% of the length of the longest genes in the file. The second requirement was that both representative of a genus were in the same group, either above or below the 90% cutoff. The two outgroup species were not required to be in the same group since they are not from the same genus. An additional 58 genes were added to the working dataset, resulting in 799 genes. A list of the 799 gene names was used to grab the corresponding amino acid files.

## Gene Alignment
Gene alignments were in `R` using a ClustalW algorithm. The gene fasta files were read into `R` using the `readDNAStringSet()` function (package: Biostrings) [@Pages2020]. Alignments were done using the `msaClustalW()` function (package: msa) [@Bodenhofer2015] using default parameters and setting maximum number of iterations to 100. The aligned functions were then converted into a readble format using `msaConvert()` (package: msa)[@Bodenhofer2015] and written into a new fasta file using the `write.fasta()` function (package: seqinr) [@Charif2007]

## Distance Matrices
After alignments, model testing was done for each gene through `MEGA-CC` [@Kumar2012] for both nucleotide and amino acid sequences. In a command line terminal, `MEGA-CC` is called along with a text file listing the file pathways to all 799 genes and a MAO file detailling the analysis preferences. Default paramaters were used for model testing. This resulted in a comma-separated values (CSV) file for each gene. These CSV files were read into `R` and the model for phylogenetic analysis was extracted based on lowest Bayesian Information Criterion (BIC). A list of the unique models and the number of genes that required each model are given in Table 2 for nucleotide sequences and in Table 3 for amino acid sequences.

```{r echo = FALSE, message = FALSE, fig.cap=TRUE}
library("tidyverse")

phylo_models_NT <- read.csv("C:/Users/Kim/OneDrive/2020_3Fall/Biology_396/9_Results_NT/Phylo_Models_NT.csv", stringsAsFactors = FALSE)
PM_NT <- data.frame(table(phylo_models_NT$BM))

colnames(PM_NT) <- c("Model", "Number")

PM_NT <- mutate(PM_NT,
             Percent = round((Number / sum(Number)) * 100, digits = 2))

library("knitr")
kable(PM_NT, caption = "Table 2. The number of genes that required each phylogenetic tree model according to model testing and the lowest BIC for nucleotide sequences.")
```

```{r echo = FALSE, message = FALSE, fig.cap=TRUE}

phylo_models_AA <- read.csv("C:/Users/Kim/OneDrive/2020_3Fall/Biology_396/9_Results_AA/Phylo_Models_AA.csv", stringsAsFactors = FALSE)
PM_AA <- data.frame(table(phylo_models_AA$BM))

colnames(PM_AA) <- c("Model", "Number")

PM_AA <- mutate(PM_AA,
                Percent = round((Number / sum(Number)) * 100, digits = 2))

library("knitr")
kable(PM_AA, caption = "Table 3. The number of genes that required each phylogenetic tree model according to model testing and the lowest BIC for amino acid sequences.")
```

For each unique model, a text file was written listing the gene file pathways of each gene that required that model. In a command line terminal, `MEGA-CC` [@Kumar2012] was called with a text file listing gene file pathways for a certain model and the corresponding MAO file. This was done for each unique model. These MAO files specified the use of a maximum likelihood statistical method, 500 bootstrap replications, the model, and rate patterns to create a phylogenetic tree for each gene. If a model required a Gamma distribution, the number of discrete Gamma categories was set to 5, which is default.

Distance matrices were extracted from each of the phylogenetic trees and saved as Excel files. These files were read into `R` and the genetic distances between both of the *Mixta* species and all of the sequences were extracted for further analysis in `R`. Tables were created in `R Markdown` using the `kable()` function (package: knitr) [@Xie2020] and figures were created using the `ggplot()` function (package: ggplot2) [@Wickham2016].

# Results and Discussion

# Conclusion

\newpage
# Literature Cited
<div id="refs"></div>

\newpage
# Appendix